---
title: "Week 7: Basis Functions and Splines"
subtitle: "*DSAN 5300: Statistical Learning*<br><span class='subsubtitle'>Spring 2025, Georgetown University</span>"
author: "Jeff Jacobs"
institute: "[`jj1088@georgetown.edu`](mailto:jj1088@georgetown.edu)"
bibliography: "../_DSAN5300.bib"
date: 2025-02-24
date-format: full
lecnum: 6
categories:
  - "Class Sessions"
format:
  revealjs:
    df-print: kable
    footer: "DSAN 5300-01 Week 7: {{< var w07.footer >}}"
    output-file: "slides.html"
    html-math-method: mathjax
    scrollable: true
    theme: [default, "../dsan-globals/jjquarto.scss"]
    slide-number: true
    echo: true
    code-fold: true
    link-external-icon: true
    link-external-newwindow: true
    include-in-header:
      text: "<link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.11.3/font/bootstrap-icons.min.css'><link rel='preconnect' href='https://fonts.googleapis.com'><link rel='preconnect' href='https://fonts.gstatic.com' crossorigin><link href='https://fonts.googleapis.com/css2?family=Kablammo&family=Rubik+Puddles&display=swap' rel='stylesheet'><link href='https://fonts.googleapis.com/css2?family=Workbench&display=swap' rel='stylesheet'><link rel='stylesheet' type='text/css' href='https://cdn.jsdelivr.net/gh/dreampulse/computer-modern-web-font@master/fonts.css'>"
    simplemenu:
      flat: true
      barhtml:
        header: "<div class='menubar'><span style='position: absolute; left: 8; padding-left: 8px;'><a href='./index.html'>&larr; Return to Notes</a></span><ul class='menu'></ul></div>"
      scale: 0.5
    revealjs-plugins:
      - simplemenu
  html:
    df-print: kable
    output-file: "index.html"
    html-math-method: mathjax
    echo: true
    code-fold: true
---

::: {.content-visible unless-format="revealjs"}

<center class='mb-3'>
<a class="h2" href="./slides.html" target="_blank">Open slides in new tab &rarr;</a>
</center>

:::

# Schedule {.smaller .small-title .crunch-title .crunch-callout data-name="Schedule"}

Today's Planned Schedule:

| | Start | End | Topic |
|:- |:- |:- |:- |
| **Lecture** | 6:30pm | 7:00pm | [Roadmap: Even More Wiggly "Linearity" &rarr;](#roadmap) |
| | 7:00pm | 7:20pm | ["Manual" Model Selection: Subsets &rarr;](#model-selection) | 
| | 7:20pm | 7:40pm | [Key Regularization Building Block: $L^p$ Norm &rarr;](#regularization-for-automatic-model-selection) |
| | 7:40pm | 8:00pm | [Regularized Regression Intro &rarr;](#regularized-regression-(finally)) |
| **Break!** | 8:00pm | 8:10pm | |
| | 8:10pm | 8:50pm | [Basically Lasso is the Coolest Thing Ever &rarr;](#the-key-plot) |
| | 8:50pm | 9:00pm | [Scary-Looking But Actually-Fun W07 Preview &rarr;](#week-7-preview-linear.noodle-jj-functions) |

: {tbl-colwidths="[12,12,12,64]"}

::: {.hidden}

```{r}
#| label: r-source-globals
source("../dsan-globals/_globals.r")
set.seed(5300)
```

:::

{{< include ../dsan-globals/_globals-tex.qmd >}}

# Roadmap {.smaller .title-10 data-stack-name="Roadmap"}

```{=html}
<style>
@font-face {
    font-family: Noodle;
    src: url('/assets/Kbnoodlemonster.ttf');
}
.workbench-jj {
  font-family: "Workbench", serif;
  /* font-optical-sizing: auto; */
  font-weight: 400;
  font-style: normal;
  font-size: 48pt !important;
}
.noodle-jj {
    font-family: "Noodle", serif;
    font-size: 52pt !important;
}
.kablammo-jj {
  font-family: "Kablammo", serif;
  font-optical-sizing: auto;
  font-weight: 400;
  font-style: normal;
  font-variation-settings:
    "MORF" 0;
  font-size: 48pt !important;
}
.rubik-jj {
  font-family: "Rubik Puddles", serif;
  font-weight: 400;
  font-style: normal;
  font-size: 48pt !important;
}
</style>
```

<center>

**[Weeks 4-6]**

</center>

:::: {.columns}
::: {.column width="50%"}

::: {#fig-noodle}

<center>

*Polynomial Regression*

[Linear Models]{.noodle-jj}

</center>

[Font Credit](http://keepinitkoolinkinderland.blogspot.com/p/fonts.html)
:::

:::
::: {.column width="50%"}

::: {#fig-workbench}

<center>

*Piecewise Regression*

[Linear Models]{.workbench-jj}

</center>

[Font Credit](https://fonts.google.com/specimen/Workbench)
:::

:::
::::

<center>

**[Today]**

</center>

:::: {.columns}
::: {.column width="50%"}

::: {#fig-kablammo}

<center>

*Splines (Truncated Power Bases)*

[Linear Models]{.kablammo-jj}

</center>

[Font Credit](https://fonts.google.com/specimen/Kablammo)
:::

:::
::: {.column width="50%"}

::: {#fig-rubik}

<center>

*Natural Splines*

[Linear Models]{.rubik-jj}

</center>

[Font Credit](https://fonts.google.com/specimen/Rubik+Puddles)
:::

:::
::::

## What Problem Are We Solving? {.smaller}

* From W04-06, you are now really good at thinking through issues of **non-linearity** in terms of **polynomial regression**
    * (Hence, why I kept "plugging in" **degree of polynomial** as our measure of **complexity**)
* Now we want to move towards handling **any kind of non-linearity**
    * (Hence, why I kept reminding you how **degree of polynomial** is *only one possible measure* of **complexity**)

## What's So Bad About Polynomial Regression? {.smaller .title-11 .crunch-title .crunch-ul .crunch-quarto-figure .crunch-li-8 .inline-90 .math-90 .crunch-p}

* Mathematically, [Weierstrass Approximation Theorem](https://en.wikipedia.org/wiki/Weierstrass_approximation_theorem) says we can model **any function** as (possibly infinite) sum of polynomials
* In practice, this can be a **horrifically bad** way to actually model things:

:::: {.columns .text-90}
::: {.column width="32%"}

<i class='bi bi-1-circle'></i> Raw data: $y = \sin(x) + \varepsilon$

```{r}
#| label: sine-data-plot
library(tidyverse) |> suppressPackageStartupMessages()
library(latex2exp) |> suppressPackageStartupMessages()
N <- 200
x_vals <- seq(from=-10, to=10, length.out=N)
y_raw <- sin(x_vals)
y_noise <- rnorm(length(y_raw), mean=0, sd=0.075)
y_vals <- y_raw + y_noise
dgp_label <- TeX("Raw Data")
data_df <- tibble(x=x_vals, y=y_vals)
base_plot <- data_df |> ggplot(aes(x=x, y=y)) +
  geom_point() +
  theme_dsan(base_size=28)
base_plot + labs(title=dgp_label)
```

<i class='bi bi-2-circle'></i> Bad (quadratic) model

```{r}
#| label: sine-poly2-fit
quad_model <- lm(y ~ poly(x,2), data=data_df)
quad_rss <- round(get_rss(quad_model), 3)
poly_label_2 <- TeX(paste0("2 $\\beta$ Coefficients: RSS = ",quad_rss))
base_plot +
  geom_smooth(method='lm', formula=y ~ poly(x,2)) +
  labs(title=poly_label_2)
```

:::
::: {.column width="2%"}

&nbsp;

:::
::: {.column width="32%"}

<i class='bi bi-3-circle'></i> Making it "better" with more complex polynomials

```{r}
#| label: sine-poly5-fit
poly5_model <- lm(y ~ poly(x,5), data=data_df)
poly5_rss <- round(get_rss(poly5_model), 3)
poly5_label <- TeX(paste0("5 $\\beta$ Coefficients: RSS = ",poly5_rss))
base_plot +
  geom_smooth(method='lm', formula=y ~ poly(x,5)) +
  labs(title=poly5_label)
```

```{r}
#| label: sine-poly10-fit
poly8_model <- lm(y ~ poly(x,8), data=data_df)
poly8_rss <- round(get_rss(poly8_model), 3)
poly8_label <- TeX(paste0("8 $\\beta$ Coefficients: RSS = ",poly8_rss))
base_plot +
  geom_smooth(method='lm', formula=y ~ poly(x,8)) +
  labs(title=poly8_label)
```

:::
::: {.column width="2%"}

&nbsp;

:::
::: {.column width="32%"}

<i class='bi bi-4-circle'></i> Using *all* data to estimate **single** parameter, by using the "correct" **basis function**!

$$
Y = \beta_0 + \beta_1 \sin(x)
$$

```{r}
#| label: sine-fit
sine_model <- lm(y ~ sin(x), data=data_df)
sine_rss <- round(get_rss(sine_model), 3)
sine_label <- TeX(paste0("Single sin(x) Coefficient: RSS = ",sine_rss))
base_plot +
  geom_smooth(method='lm', formula=y ~ sin(x)) +
  labs(title=sine_label)
```

:::
::::

# Basis Functions {data-stack-name="Basis Functions"}

* Seems like it'll help to have a toolkit filled with different **basis functions**, so we can use them with `lm()` 🤔
* ...Let's get into it!

## The General Form {.crunch-title .crunch-ul .inline-90 .math-90}

*(Decomposing Fancy Regressions into Core "Pieces")*

* Q: What do **all** these types of regression have in common?
* A: They can all be written in the form
  
  $$
  Y = \beta_0 + \beta_1 b_1(X) + \beta_2 b_2(X) + \cdots + \beta_d b_d(X)
  $$

* Where $b(\cdot)$ is called a **basis function**
  * Polynomial: $b_j(X) = X^j$
  * Piecewise: $b_j(X) = \mathbb{1}[c_{j-1} \leq X < c_j]$
  * Earlier Example: $b_1(X) = \sin(X)$

# Segments $\rightarrow$ Splines {data-stack-name="Segmented Regression" .smaller .crunch-title .table-va}

| | What we want | How we do it | Name for result |
| - | - | - |:-:|
| <i class='bi bi-1-circle'></i> | Model **sub-regions** of $\text{domain}(X)$ | Chop $\text{domain}(X)$ into pieces, one regression per piece | *Discontinuous Segmented Regression* |
| <i class='bi bi-2-circle'></i> | **Continuous** prediction function | Require pieces to **join** at chop points $\{\xi_1, \ldots, \xi_K\}$ | *Continuous Segmented Regression* |
| <i class='bi bi-3-circle'></i> | **Smooth** prediction function | Require pieces to **join** and have **equal derivatives** at $\xi_i$ | **Spline** |
| <i class='bi bi-4-circle'></i> | Less jump in variance at **boundaries** | Reduce complexity of polynomials at endpoints | **Natural Spline** |

: {tbl-colwidths="[4,28,38,30]"}

## Discontinuous Segmented Regression {.smaller .title-12}

* Here we see why our first two basis functions were **polynomial** and **piecewise**: most rudimentary spline fits a **polynomial** to each **piece**

```{r}
#| label: quad-naive-spline
library(tidyverse) |> suppressPackageStartupMessages()
set.seed(5300)
compute_y <- function(x) {
  return(x * cos(x^2))
}
N <- 500
xmin <- -1.9
xmax <- 2.7
x_vals <- runif(N, min=xmin, max=xmax)
y_raw = compute_y(x_vals)
y_noise = rnorm(N, mean=0, sd=0.5)
y_vals <- y_raw + y_noise
prod_df <- tibble(x=x_vals, y=y_vals)
knot <- (xmin + xmax) / 2
prod_df <- prod_df |> mutate(segment = x <= knot)
# First segment model
#left_df <- prod_df |> filter(x <= knot_point)
#left_model <- lm(y ~ poly(x, 2), data=left_df)
prod_df |> ggplot(aes(x=x, y=y, group=segment)) +
  geom_point(size=0.5) +
  geom_vline(xintercept=knot, linetype="dashed") +
  geom_smooth(method='lm', formula=y ~ poly(x, 2), se=TRUE) +
  theme_classic(base_size=22)
```

* What's the issue with this?

## Forcing Continuity at Knot Points {.smaller .title-12}

* Starting slow: let's come back down to line world...
* Why are these two lines "allowed" to be non-continuous, under our current approach?

:::: {.columns}
::: {.column width="50%"}

$$
\hspace{6.5cm} Y^{\phantom{🧐}}_L = \beta_0 + \beta_1 X_L
$$

:::
::: {.column width="50%"}

$$
Y_R = \beta_0^{🧐} + \beta_1 X_R \hspace{5cm}
$$

:::
::::

```{r}
#| label: linear-continuity
set.seed(5300)
xmin_sub <- -1
xmax_sub <- 1.9
sub_df <- prod_df |> filter(x >= xmin_sub & x <= xmax_sub)
sub_knot <- (xmin_sub + xmax_sub) / 2
sub_df <- sub_df |> mutate(segment = x <= sub_knot)
# First segment model
sub_df |> ggplot(aes(x=x, y=y, group=segment)) +
  geom_point(size=0.5) +
  geom_vline(xintercept=sub_knot, linetype="dashed") +
  geom_smooth(method='lm', formula=y ~ x, se=TRUE, linewidth=g_linewidth) +
  theme_classic(base_size=28)
```

* ...We need $X_R$ to have its own slope but **not** its own intercept! Something like:

$$
Y = \beta_0 + \beta_1 (X \text{ before } \xi) + \beta_2 (X \text{ after } \xi)
$$

## Truncated Power Basis (ReLU Basis) {.crunch-title .math-80 .inline-90 .title-09}

$$
\text{ReLU}(x) \definedas (x)_+ \definedas \begin{cases}
0 &\text{if }x \leq 0 \\
x &\text{if }x > 0
\end{cases} \implies (x - \xi)_+ = \begin{cases}
0 &\text{if }x \leq \xi \\
x - \xi &\text{if }x > 0
\end{cases}
$$

```{r}
#| label: truncated-power
library(latex2exp) |> suppressPackageStartupMessages()
trunc_x <- function(x, xi) {
  return(ifelse(x <= xi, 0, x - xi))
}
trunc_x_05 <- function(x) trunc_x(x, 1/2)
trunc_title <- TeX("$(x - \\xi)_+$ with $\\xi = 0.5$")
trunc_label <- TeX("$y = (x - \\xi)_+$")
ggplot() +
  stat_function(
    data=data.frame(x=c(-3,3)),
    fun=trunc_x_05,
    linewidth=g_linewidth
  ) +
  xlim(-3, 3) +
  theme_dsan(base_size=28) +
  labs(
    title = trunc_title,
    x = "x",
    y = trunc_label
  )
```

* $\Rightarrow$ we can include a **"slope modifier"** $\beta_m (X - \xi)_+$ that only **"kicks in"** once $x$ goes past $\xi$! (Changing slope by $\beta_m$)

## Linear Segments with Continuity at Knot {.smaller .title-12 .math-80 .table-80}

:::: {.columns}
::: {.column width="40%"}

Our new (non-naïve) model:

$$
\begin{align*}
Y &= \beta_0 + \beta_1 X + \beta_2 (X - \xi)_+ \\
&= \beta_0 + \begin{cases}
\beta_1 X &\text{if }X \leq \xi \\
(\beta_1 + \beta_2)X &\text{if }X > \xi
\end{cases}
\end{align*}
$$

:::
::: {.column width="60%"}

```{r}
#| label: linear-seg-model
sub_df <- sub_df |> mutate(x_tr = ifelse(x < sub_knot, 0, x - sub_knot))
linseg_model <- lm(y ~ x + x_tr, data=sub_df)
broom::tidy(linseg_model) |> mutate_if(is.numeric, round, 3) |> select(-p.value)
```

:::
::::

```{r}
#| label: linear-seg-plot
sub_df |> ggplot(aes(x=x, y=y)) +
  geom_point(size=0.5) +
  geom_vline(xintercept=sub_knot, linetype="dashed") +
  geom_smooth(method='lm', formula=y ~ x + ifelse(x > sub_knot, x-sub_knot, 0), se=TRUE, linewidth=g_linewidth) +
  theme_classic(base_size=28)
```

## Continuous Segmented Regression {.smaller .crunch-ul .crunch-title .title-12 .math-80 .table-80}

:::: {.columns}
::: {.column width="50%"}

$$
\begin{align*}
Y &= \beta_0 + \beta_1 X + \beta_2 X^2 + \beta_3 (X - \xi)_+ + \beta_4 (X - \xi)_+^2 \\[0.8em]
&= \beta_0 + \begin{cases}
\beta_1 X + \beta_2 X^2 &\text{if }X \leq \xi \\
(\beta_1 + \beta_3) X + (\beta_2 + \beta_4) X^2 &\text{if }X > \xi
\end{cases}
\end{align*}
$$

:::
::: {.column width="50%"}

```{r}
#| label: continuous-seg-model
#| code-fold: show
prod_df <- prod_df |> mutate(x_tr = ifelse(x > knot, x - knot, 0))
seg_model <- lm(
  y ~ poly(x, 2) + poly(x_tr, 2),
  data=prod_df
)
```

:::
::::

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: continuous-seg-model-results
seg_model |> broom::tidy() |> mutate_if(is.numeric, round, 3) |> select(-p.value)
```

:::
::: {.column width="50%"}

```{r}
#| label: continuous-seg-plot
cont_seg_plot <- ggplot() +
  geom_point(data=prod_df, aes(x=x, y=y), size=0.5) +
  geom_vline(xintercept=knot, linetype="dashed") +
  stat_smooth(
    data=prod_df, aes(x=x, y=y),
    method='lm',
    formula=y ~ poly(x,2) + poly(ifelse(x > knot, (x - knot), 0), 2),
    n = 300
  ) +
  #geom_smooth(data=prod_df |> filter(segment == FALSE), aes(x=x, y=y), method='lm', formula=y ~ poly(x,2)) +
  # geom_smooth(method=segreg, formula=y ~ seg(x, npsi=1, fixed.psi=0.5)) + # + seg(I(x^2), npsi=1)) +
  theme_classic(base_size=22)
cont_seg_plot
```

:::
::::

* There's still a problem here... can you see what it is? (Hint: things that break calculus)

# Splines (The Moment You've Been Waiting For!) {data-stack-name="Splines" .title-12}

## Constrained Derivatives: "Quadratic Splines" (⚠️) {.smaller .crunch-title .title-10 .crunch-quarto-figure .crunch-math}

* Like "Linear Probability Models", "Quadratic Splines" are a **red flag**
* Why? If we have leftmost plot below, but want it differentiable everywhere... only option is to fit a single quadratic, defeating the whole point of the "chopping"!

$$
\frac{\partial Y}{\partial X} = \beta_1 + 2\beta_2 X + \beta_3 + 2\beta_4 X
$$

:::: {.columns}
::: {.column width="30%"}

Continuous Segmented:

```{r}
#| label: quadratic-seg-plot-again
cont_seg_plot
```

:::
::: {.column width="5%"}

&nbsp;

:::
::: {.column width="30%"}

"Quadratic Spline":

```{r}
#| label: deriv-contraints
ggplot() +
  geom_point(data=prod_df, aes(x=x, y=y), size=0.5) +
  geom_vline(xintercept=knot, linetype="dashed") +
  stat_smooth(
    data=prod_df, aes(x=x, y=y),
    method='lm',
    formula=y ~ poly(x,2) + ifelse(x > knot, (x - knot)^2, 0),
    n = 300
  ) +
  theme_dsan(base_size=22)
```

:::
::: {.column width="5%"}

&nbsp;

:::
::: {.column width="30%"}

$$
Y = \beta_0 + \beta_1 X + \beta_2 X^2
$$

```{r}
#| label: base-quadratic-reg
ggplot() +
  geom_point(data=prod_df, aes(x=x, y=y), size=0.5) +
  geom_vline(xintercept=knot, linetype="dashed") +
  stat_smooth(
    data=prod_df, aes(x=x, y=y),
    method='lm',
    formula=y ~ poly(x,2),
    n = 300
  ) +
  theme_dsan(base_size=22)
```

:::
::::

* $\implies$ Least-complex function that allows smooth "joining" is **cubic function**

## Cubic Splines (aka, Splines) {.smaller .crunch-title .crunch-ul}

*(We did it, we finally did it)*

```{r}
#| label: cubic-spline-plot
ggplot() +
  geom_point(data=prod_df, aes(x=x, y=y), size=0.5) +
  geom_vline(xintercept=knot, linetype="dashed") +
  stat_smooth(
    data=prod_df, aes(x=x, y=y),
    method='lm',
    formula=y ~ poly(x,3) + ifelse(x > knot, (x - knot)^3, 0),
    n = 300
  ) +
  theme_dsan(base_size=22)
```

## Why Not Stop There? {.smaller .crunch-title .crunch-ul}

:::: {.columns}
::: {.column width="50%"}

Polynomials start to BEHAVE BADLY as they go to $-\infty$ and $\infty$

```{r}
#| label: cubic-spline-boundary-issue
library(splines) |> suppressPackageStartupMessages()
set.seed(5300)
N_sparse <- 400
x_vals <- runif(N_sparse, min=xmin, max=xmax)
y_raw = compute_y(x_vals)
y_noise = rnorm(N_sparse, mean=0, sd=1.0)
y_vals <- y_raw + y_noise
sparse_df <- tibble(x=x_vals, y=y_vals)
knot_sparse <- (xmin + xmax) / 2
ggplot() +
  geom_point(data=sparse_df, aes(x=x, y=y), size=0.5) +
  geom_vline(xintercept=knot, linetype="dashed") +
  stat_smooth(
    data=sparse_df, aes(x=x, y=y),
    method='lm',
    formula=y ~ bs(x, knots=c(knot_sparse), degree=25),
    n=300
  ) +
  theme_dsan(base_size=22)
```

:::
::: {.column width="50%"}

Natural Splines: **Force** leftmost and rightmost pieces to be **linear**

```{r}
#| label: natural-spline
library(splines) |> suppressPackageStartupMessages()
set.seed(5300)
N_sparse <- 400
x_vals <- runif(N_sparse, min=xmin, max=xmax)
y_raw = compute_y(x_vals)
y_noise = rnorm(N_sparse, mean=0, sd=1.0)
y_vals <- y_raw + y_noise
sparse_df <- tibble(x=x_vals, y=y_vals)
knot_sparse <- (xmin + xmax) / 2
ggplot() +
  geom_point(data=sparse_df, aes(x=x, y=y), size=0.5) +
  geom_vline(xintercept=knot, linetype="dashed") +
  stat_smooth(
    data=sparse_df, aes(x=x, y=y),
    method='lm',
    formula=y ~ ns(x, knots=c(-1.8,-1.7,-1.6,-1.5,-1.4,-1.2,-1.0,-0.8,-0.6,-0.4,-0.2,0,knot_sparse,1,1.5,2), Boundary.knots=c(xmin + 0.1, xmax - 0.1)),
    n=300
  ) +
  theme_dsan(base_size=22)
```

:::
::::

## References

::: {#refs}
:::


