---
title: "Week 8: Support Vector Machines"
subtitle: "*DSAN 5300: Statistical Learning*<br><span class='subsubtitle'>Spring 2025, Georgetown University</span>"
author: "Jeff Jacobs"
institute: "[`jj1088@georgetown.edu`](mailto:jj1088@georgetown.edu)"
bibliography: "../_DSAN5300.bib"
date: 2025-03-10
date-format: full
lecnum: 8
categories:
  - "Class Sessions"
format:
  revealjs:
    df-print: kable
    footer: "DSAN 5300-01 Week 8: {{< var w08.footer >}}"
    output-file: "slides.html"
    html-math-method: mathjax
    scrollable: true
    theme: [default, "../dsan-globals/jjquarto.scss"]
    slide-number: true
    echo: true
    code-fold: true
    link-external-icon: true
    link-external-newwindow: true
    include-in-header:
      text: "<link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.11.3/font/bootstrap-icons.min.css'><link rel='stylesheet' type='text/css' href='https://cdn.jsdelivr.net/gh/dreampulse/computer-modern-web-font@master/fonts.css'>"
    simplemenu:
      flat: true
      barhtml:
        header: "<div class='menubar'><span style='position: absolute; left: 8; padding-left: 8px;'><a href='./index.html'>&larr; Return to Notes</a></span><ul class='menu'></ul></div>"
      scale: 0.5
    revealjs-plugins:
      - simplemenu
  html:
    df-print: kable
    output-file: "index.html"
    html-math-method: mathjax
    echo: true
    code-fold: true
---

::: {.content-visible unless-format="revealjs"}

<center class='mb-3'>
<a class="h2" href="./slides.html" target="_blank">Open slides in new tab &rarr;</a>
</center>

:::

# Schedule {.smaller .small-title .crunch-title .crunch-callout data-name="Schedule"}

Today's Planned Schedule:

| | Start | End | Topic |
|:- |:- |:- |:- |
| **Lecture** | 6:30pm | 7:00pm | [Separating Hyperplanes &rarr;](#learning-decision-boundaries) |
| | 7:00pm | 7:20pm | [Max-Margin Classifier &rarr;](#the-max-margin-classifier) | 
| | 7:20pm | 8:00pm | [Support Vector *Classifiers* &rarr;](#support-vector-classifiers) |
| **Break!** | 8:00pm | 8:10pm | |
| | 8:10pm | 9:00pm | [Support Vector *Machines* &rarr;](#support-vector-machines) |

: {tbl-colwidths="[12,12,12,64]"}

::: {.hidden}

```{r}
#| label: r-source-globals
source("../dsan-globals/_globals.r")
set.seed(5300)
```

:::

{{< include ../dsan-globals/_globals-tex.qmd >}}

## Quick Roadmap

* Weeks 1-7: **Regression** was central motivating problem
  * (Even when we looked at classification *via* Logistic Regression, it was regression and then a final **thresholding** step on the regression result)
* This Week:
  * On the one hand, we shift focus to **binary classification** through learning **decision boundaries**
  * On the other hand, **everything we've discussed about handling non-linearity** will come into play!

# Max-Margin Classifiers {data-stack-name="Max-Margin Classifiers"}

* Most important **building block** for SVMs...
* But should **not** be used on its own! For reasons we'll see shortly
  * (Like "Linear Probability Models" or "Quadratic Splines", MMCs highlight problem that [usable method] solves!)
* MMC $\leadsto$ Support Vector Classifiers (can be used)
* MMC $\leadsto$ Support Vector Machines (can be used)

## Learning Decision Boundaries {.title-09 .crunch-title .crunch-quarto-figure .crunch-ul .crunch-li-8 .crunch-p}

:::: {.columns}
::: {.column width="50%"}

* How can we get our computer to "learn" which houses I like?

```{r}
#| label: sup-data-plot
#| crop: false
#| fig-width: 4.5
#| fig-height: 4.5
library(tidyverse)
house_df <- tibble::tribble(
  ~sqm, ~yrs, ~Rating,
  10, 5, "Disliked",
  10, 20, "Disliked",
  20, 22, "Liked",
  25, 12, "Liked",
  11, 15, "Disliked",
  22.1, 5, "Liked"
) |> mutate(
  label = ifelse(Rating == "Liked", 1, -1)
)
base_plot <- house_df |> ggplot(aes(x=sqm, y=yrs)) +
  labs(
    title = "Jeff's House Search",
    x = "Square Meters",
    y = "Years Old"
  ) +
  expand_limits(x=c(0,25), y=c(0,25)) +
  coord_equal() +
  # 45 is minus sign, 95 is em-dash
  scale_shape_manual(values=c(95, 43)) +
  theme_dsan(base_size=14)
base_plot +
  geom_point(
    aes(color=Rating, shape=Rating), size=g_pointsize * 0.9,
    stroke=6
  ) +
  geom_point(aes(fill=Rating), color='black', shape=21, size=6, stroke=0.75, alpha=0.333)
```

:::
::: {.column width="50%"}

* Naïve idea: Try random lines (each forming a **decision boundary**), pick "best" one

```{r}
#| label: random-lines-plot
#| crop: false
#| fig-width: 4.5
#| fig-height: 4.5
set.seed(5300)
is_separating <- function(beta_vec) {
  beta_str <- paste0(beta_vec, collapse=",")
  # print(paste0("is_separating: ",beta_str))
  margins <- c()
  for (i in 1:nrow(house_df)) {
    cur_data <- house_df[i,]
    # print(cur_data)
    linear_comb <- beta_vec[1] + beta_vec[2] * cur_data$sqm + beta_vec[3] * cur_data$yrs
    cur_margin <- cur_data$label * linear_comb
    # print(cur_margin)
    margins <- c(margins, cur_margin)
  }
  #print(margins)
  return(all(margins > 0) | all(margins < 0))
}
cust_rand_lines_df <- tribble(
  ~b0, ~b1, ~b2,
  # 41, -0.025, -1,
  165, -8, -1,
  -980, 62, -1
) |> mutate(
  slope=-(b1/b2),
  intercept=-(b0/b2)
)
num_lines <- 20
rand_b0 <- runif(num_lines, min=-40, max=40)
rand_b1 <- runif(num_lines, min=-2, max=2)
# rand_b2 <- -1 + 2*rbernoulli(num_lines)
rand_b2 <- -1
rand_lines_df <- tibble::tibble(
  id=1:num_lines,
  b0=rand_b0,
  b1=rand_b1,
  b2=rand_b2
) |> mutate(
  slope=-(b1/b2),
  intercept=-(b0/b2)
)
rand_lines_df <- bind_rows(rand_lines_df, cust_rand_lines_df)
# Old school for loop
for (i in 1:nrow(rand_lines_df)) {
  cur_line <- rand_lines_df[i,]
  cur_beta_vec <- c(cur_line$b0, cur_line$b1, cur_line$b2)
  cur_is_sep <- is_separating(cur_beta_vec)
  rand_lines_df[i, "is_sep"] <- cur_is_sep
}
base_plot +
  geom_abline(
    data=rand_lines_df, aes(slope=slope, intercept=intercept), linetype="dashed"
  ) +
  geom_point(
    aes(color=Rating, shape=Rating), size=g_pointsize * 0.9,
    stroke=6
  ) +
  geom_point(aes(fill=Rating), color='black', shape=21, size=6, stroke=0.75, alpha=0.333) +
  labs(
    title = paste0("10 Boundary Guesses"),
    x = "Square Meters",
    y = "Years Old"
  )
```

<center>

[*(...Which one is "best"?)*]{style="font-size: 75%"}

</center>

:::
::::

## Separating Hyperplanes {.crunch-title .title-09 .crunch-quarto-figure .crunch-ul}

* Any line which **perfectly separates** positive from negative cases is a **separating hyperplane**

```{r}
#| label: sep-lines-plot
#| crop: false
base_plot +
  geom_abline(
    data=rand_lines_df, aes(slope=slope, intercept=intercept, linetype=is_sep)
  ) +
  geom_abline(
    data=rand_lines_df |> filter(is_sep),
    aes(slope=slope, intercept=intercept),
    linewidth=3, color=cb_palette[4], alpha=0.333
  ) +
  scale_linetype_manual("Separating?", values=c("dotted", "dashed")) +
  geom_point(
    aes(color=Rating, shape=Rating), size=g_pointsize * 0.9,
    stroke=6
  ) +
  geom_point(aes(fill=Rating), color='black', shape=21, size=6, stroke=0.75, alpha=0.333) +
  labs(
    title = paste0("The Like vs. Dislike Boundary: 10 Guesses"),
    x = "Square Meters",
    y = "Years Old"
  )
```

* Why might the **left** line be better? *(Think of DGP!)*

## Max-Margin Hyperplanes {.crunch-title .title-11 .smaller .crunch-quarto-figure .crunch-ul .crunch-li-8}

* Among the two, left hyperplane establishes a larger **margin** between the two classes!
* **Margin** (of hyperplane) = **Smallest distance** to datapoint (Think of hyperplane as middle of separating **slab**: how wide is the slab?)
* **Important**: Note how margin is **not** the same as *average* distance!

:::: {.columns}
::: {.column width="50%"}

```{r}
#| label: left-sep-only-plot
#| crop: false
sep_lines_df <- rand_lines_df |> filter(is_sep) |> mutate(
  norm_slope = (-1)/slope
)
cur_line_df <- sep_lines_df |> filter(slope > 0)
# left_line_df
# And make one copy per point
cur_sup_df <- uncount(cur_line_df, nrow(house_df))
cur_sup_df <- bind_cols(cur_sup_df, house_df)
cur_sup_df <- cur_sup_df |> mutate(
  norm_intercept = yrs - norm_slope * sqm,
  margin_intercept = yrs - slope * sqm,
  margin_intercept_gap = intercept - margin_intercept,
  margin_intercept_inv = intercept + margin_intercept_gap,
  norm_cross_x = -(norm_intercept - intercept) / (norm_slope - slope),
  x_gap = norm_cross_x - sqm,
  norm_cross_y = yrs + x_gap * norm_slope,
  vec_margin = label * (b0 + b1 * sqm + b2 * yrs),
  is_sv = vec_margin <= 240
)
base_plot +
  geom_abline(
    data=cur_line_df, aes(slope=slope, intercept=intercept), linetype="solid"
  ) +
  geom_abline(
    data=cur_sup_df |> filter(is_sv),
    aes(
      slope=slope,
      intercept=margin_intercept
    ),
    linetype="dashed"
  ) +
  geom_abline(
    data=cur_sup_df |> filter(is_sv),
    aes(
      slope=slope,
      intercept=margin_intercept_inv
    ),
    linetype="dashed"
  ) +
  geom_segment(
    data=cur_sup_df |> filter(is_sv),
    aes(x=sqm, y=yrs, xend=norm_cross_x, yend=norm_cross_y),
    color=cb_palette[4], linewidth=3
  ) +
  geom_segment(
    data=cur_sup_df,
    aes(x=sqm, y=yrs, xend=norm_cross_x, yend=norm_cross_y, linetype=is_sv)
  ) +
  geom_point(
    aes(color=Rating, shape=Rating), size=g_pointsize * 0.9,
    stroke=6
  ) +
  geom_point(aes(fill=Rating), color='black', shape=21, size=6, stroke=0.75, alpha=0.333) +
  scale_linetype_manual("Support\nVector?", values=c("dotted", "solid")) +
  labs(
    title = paste0("Left Hyperplane Distances"),
    x = "Square Meters",
    y = "Years Old"
  )
```

:::
::: {.column width="50%"}

```{r}
#| label: right-sep-only-plot
#| crop: false
# New calculation: line with same slope but that hits the SV
# y - y1 = m(x - x1), so...
# y - yrs = m(x - sqm) <=> y = m(x-sqm) + yrs <=> y = mx  - m*sqm + yrs
# <=> b = yrs - -m*sqm
cur_line_df <- sep_lines_df |> filter(slope < 0)
# left_line_df
# And make one copy per point
cur_sup_df <- uncount(cur_line_df, nrow(house_df))
cur_sup_df <- bind_cols(cur_sup_df, house_df)
cur_sup_df <- cur_sup_df |> mutate(
  norm_intercept = yrs - norm_slope * sqm,
  margin_intercept = yrs - slope * sqm,
  margin_intercept_gap = intercept - margin_intercept,
  margin_intercept_inv = intercept + margin_intercept_gap,
  norm_cross_x = -(norm_intercept - intercept) / (norm_slope - slope),
  x_gap = norm_cross_x - sqm,
  norm_cross_y = yrs + x_gap * norm_slope,
  vec_margin = abs(label * (b0 + b1 * sqm + b2 * yrs)),
  is_sv = vec_margin <= 25
)
base_plot +
  geom_abline(
    data=cur_line_df, aes(slope=slope, intercept=intercept), linetype="solid"
  ) +
  geom_abline(
    data=cur_sup_df |> filter(is_sv),
    aes(
      slope=slope,
      intercept=margin_intercept
    ),
    linetype="dashed"
  ) +
  geom_abline(
    data=cur_sup_df |> filter(is_sv),
    aes(
      slope=slope,
      intercept=margin_intercept_inv
    ),
    linetype="dashed"
  ) +
  # geom_abline(
  #   data=cur_line_df,
  #   aes(slope=slope, intercept=intercept),
  #   linewidth=3, color=cb_palette[4], alpha=0.333
  # ) +
  geom_segment(
    data=cur_sup_df |> filter(vec_margin <= 18),
    aes(x=sqm, y=yrs, xend=norm_cross_x, yend=norm_cross_y),
    color=cb_palette[4], linewidth=3
  ) +
  geom_segment(
    data=cur_sup_df,
    aes(x=sqm, y=yrs, xend=norm_cross_x, yend=norm_cross_y, linetype=is_sv)
  ) +
  geom_point(
    aes(color=Rating, shape=Rating), size=g_pointsize * 0.9,
    stroke=6
  ) +
  geom_point(aes(fill=Rating), color='black', shape=21, size=6, stroke=0.75, alpha=0.333) +
  scale_linetype_manual("Support\nVector?", values=c("dotted", "solid")) +
  labs(
    title = paste0("Right Hyperplane Margin"),
    x = "Square Meters",
    y = "Years Old"
  )
```

:::
::::

## *Optimal* Max-Margin Hyperplane {.crunch-title .smaller}

* On previous slides we chose from a small set of *randomly-generated* lines...
* Now that we have our **max-margin objective**, we can optimize over **all possible lines!**

:::: {.columns}
::: {.column width="50%"}

$$
\begin{align*}
\text{maximize } & M \\
\text{s.t. } & y_i(\beta_0 + \beta_1x_i) > 0
\end{align*}
$$

:::
::: {.column width="50%"}

```{r}
#| label: optimal-mmh
#| crop: false
#| fig-width: 5
#| fig-height: 5
library(e1071)
liked <- as.factor(house_df$Rating == "Liked")
cent_df <- house_df
cent_df$sqm <- scale(cent_df$sqm)
cent_df$yrs <- scale(cent_df$yrs)
svm_model <- svm(liked ~ sqm + yrs, data=cent_df, kernel="linear")
cf <- coef(svm_model)
sep_intercept <- -cf[1] / cf[3]
sep_slope <- -cf[2] / cf[3]
cent_df |> ggplot(aes(x=sqm, y=yrs)) +
  coord_equal() +
  scale_shape_manual(values=c(95, 43)) +
  theme_dsan(base_size=14) +
  geom_abline(
    intercept=sep_intercept, slope=sep_slope, linetype="solid"
  ) +
  # geom_abline(
  #   data=cur_line_df,
  #   aes(slope=slope, intercept=intercept),
  #   linewidth=3, color=cb_palette[4], alpha=0.333
  # ) +
  geom_point(
    aes(color=Rating, shape=Rating), size=g_pointsize * 0.9,
    stroke=6
  ) +
  geom_point(aes(fill=Rating), color='black', shape=21, size=6, stroke=0.75, alpha=0.333) +
  scale_linetype_manual("Support\nVector?", values=c("dotted", "solid")) +
  labs(
    title = "Optimal Max-Margin Hyperplane",
    x = "Z(Square Meters)",
    y = "Z(Years Old)"
  )
```

:::
::::

## When Does This Fail?

* Here we need to keep in mind how our goal is relative to the **Data-Generating Process!**

# Support Vector Classifiers {data-stack-name="SVCs"}

## Handling Non-Linearly-Separable Data

# Support Vector *Machines* {data-stack-name="SVMs"}

## Non-Linearity Strikes Again!

